---
title: "Case Study: Divvy Bike share"
output: html_document
---

#### This is a case study using the fictitious company Cyclistic, established in 2016 and based in Chicago. : A bike-share program that features more than 5,800 bicycles and 600 docking stations. Cyclistic sets itself apart by also offering reclining bikes, hand tricycles, and cargo bikes, making bike-share more inclusive to people with disabilities and riders who can't use a standard two-wheeled bike. The majority of riders opt for traditional bikes; about 8% of riders use the assistive options. Cyclistic users are more likely to ride for leisure, but about 30% use them to commute to work each day.

## Key Stakeholders

-   **Lily Moreno:** The director of marketing and my manager. Responsible for the development of campaigns and initiatives to promote the bike-share program. These may include email, social media, and other channels

-   **Cyclistic exectutive team**:The notoriously detail-oriented executive team will decide whether to approve the recommended marketing program

-   **Cyclistic marketing analytics team:** A team of data analysts who are responsible for collecting, analyzing, and reporting data that helps guide Cyclistic marketing strategy

## Summary

After the Cylistic's finance analyst concluded that annual members are more profitable than casual riders. Our director of marketing, Lily Moreno believes the future success of the company depends on maximizing the number of annual memberships.

Moreno has set a clear goal: Design marketing strategies aimed at converting casual riders into annual members. In order to do that, however, the marketing analyst team needs to better understand how annual members and casual riders differ, why casual riders would buy a membership, and how digital media could affect their marketing tactics. Moreno and her team are interested in analyzing the Cyclistic's historical bike trip data to identify trends. Rather than creating a marketing campaign that targets all-new customers, Moreno believes there is a very good chance to convert casual riders into members

## **Task**

As a junior data analyst working with the marketing analyst team. I have been tasked with finding out how annual members and casual riders use bikes differently. The team wants to understand how casual riders and annual members use Cyclistic bikes differently. From these findings,the team will design a new marketing strategy to convert casual riders into annual members. But first, Cyclistic executives must approve your recommendations, so they must be backed up with compelling data insights and professional data visualizations

## Data Used

###### Will use Cyclistic's historical trip data to analyze and identify trends.Note, the datasets have a different name because Cyclistic is a fictional company. For the purposes of this case study, the datasets are appropriate and will enable me to answer the business questions. The data has been made available by Motivate International Inc. under this license <https://www.divvybikes.com/data-license-agreement> This is public data that you can use to explore how different customer types are using Cyclistic bikes. But note that data-privacy issues prohibit you from using riders' personally identifiable information. This means that you won't be able to connect pass purchases to credit card numbers to determine if casual riders live in the Cyclistic service area or if they have purchased multiple single passes.

##### Installed and loaded the needed packages. Tidyverse for importing and wrangling, lubridate for date functions and ggplot2 for my visualizations

```{r}
install.packages("tidyverse")
install.packages("lubridate")
install.packages("ggplot2")
```

```{r}
library(tidyverse)
library(lubridate)
library(ggplot2)
```

##### Imported data the read_csv command.Then Examined my colunm names to make sure they match so I can eventually bind all files.

```{r}
q2_2019 <- read_csv("Divvy_Trips_2019_Q2.csv")
q3_2019 <- read_csv("Divvy_Trips_2019_Q3.csv")
q4_2019 <- read_csv("Divvy_Trips_2019_Q4.csv")
q1_2020 <- read_csv("Divvy_Trips_2020_Q1.csv")

```

```{r}
colnames(q2_2019)
colnames(q3_2019)
colnames(q4_2019)
colnames(q1_2020)
```

##### Renamed Columns to make them consitent with q1_2020, which is the format the company is using moving forward. Using rename() to have all column names match.

```{r}
q2_2019 <- rename(q2_2019 
                  ,ride_id = "01 - Rental Details Rental ID"
                  ,rideable_type = "01 - Rental Details Bike ID"
                  ,started_at = "01 - Rental Details Local Start Time"
                  ,ended_at = "01 - Rental Details Local End Time"
                  ,start_station_name = "03 - Rental Start Station Name"
                  ,start_station_id = "03 - Rental Start Station ID"
                  ,end_station_name = "02 - Rental End Station Name"
                  ,end_station_id = "02 - Rental End Station ID"
                  ,member_casual = "User Type")

q3_2019 <- rename(q3_2019
                  ,ride_id = trip_id
                  ,rideable_type = bikeid 
                  ,started_at = start_time  
                  ,ended_at = end_time  
                  ,start_station_name = from_station_name 
                  ,start_station_id = from_station_id 
                  ,end_station_name = to_station_name 
                  ,end_station_id = to_station_id 
                  ,member_casual = usertype)
                  

q4_2019 <- rename(q4_2019
                   ,ride_id = trip_id
                   ,rideable_type = bikeid 
                   ,started_at = start_time  
                   ,ended_at = end_time  
                   ,start_station_name = from_station_name 
                   ,start_station_id = from_station_id 
                   ,end_station_name = to_station_name 
                   ,end_station_id = to_station_id 
                   ,member_casual = usertype)
```

##### Inspected the newly renamed data for inacurracies.

```{r}
str(q1_2020)
str(q4_2019)
str(q3_2019)
str(q2_2019)
```

##### Converted ride_id and rideable_type to character so that they can stack correctly using the mutate command.

```{r}
q2_2019 <-  mutate(q2_2019, ride_id = as.character(ride_id)
                   ,rideable_type = as.character(rideable_type)) 
q3_2019 <-  mutate(q3_2019, ride_id = as.character(ride_id)
                   ,rideable_type = as.character(rideable_type)) 
q4_2019 <-  mutate(q4_2019, ride_id = as.character(ride_id)
                   ,rideable_type = as.character(rideable_type)) 

```

##### Now that my data is all matching, I can Bind my Data into one Data Frame and remove columns I won't be using for my analysis.

```{r}
all_trips <- bind_rows(q2_2019, q3_2019, q4_2019, q1_2020)
```

```{r}
all_trips <- all_trips %>% select(-c(start_lat, start_lng, end_lat, end_lng, birthyear, gender, "01 - Rental Details Duration In Seconds Uncapped", "05 - Member Details Member Birthday Year", "Member Gender", "tripduration"))

```

## clean and prep my data for analysis

##### Inspected the new table created, all_trips

```{r}
colnames(all_trips)
nrow(all_trips)
dim(all_trips)
head(all_trips)
tail(all_trips)
str(all_trips)
summary(all_trips)
```

##### Found a few problems in that need to be fixed.

-   In the "member_casual" column, replace "Subscriber" with "member" and "Customer" with "casual" Before 2020, Divvy used different labels for these two types of riders. Will want to make our dataframe consistent with their current handle
-   The data can only be aggregated at the ride-level, which is too granular. We will want to add some additional columns of data, such as -- day, month, year,-- that provide additional opportunities to aggregate the data.
-   We will want to add a calculated field for length of ride since the 2020Q1 data did not have the "tripduration" column. We will add "ride_length" to the entire dataframe for consistency.
-   There are some rides where tripduration shows up as negative, including several hundred rides where Divvy took bikes out of circulation for Quality Control reasons. We will want to delete these rides.

##### Checking my table to see how many observations fall under each usertype and then reassign them to the desired value

```{r}
table(all_trips$member_casual)
```

```{r}
all_trips <-  all_trips %>% 
  mutate(member_casual = recode(member_casual
                           ,"Subscriber" = "member"
                           ,"Customer" = "casual"))

```

##### Verifying my new table to make sure all proper observations were reassigned

```{r}
table(all_trips$member_casual)
```

##### Add columns that list the date, month, day, and year of each ride. This will allow me to aggregate ride data for each month, day or year.

```{r}
all_trips$date <- as.Date(all_trips$started_at) #The default format is yyyy-mm-dd
all_trips$month <- format(as.Date(all_trips$date), "%m")
all_trips$day <- format(as.Date(all_trips$date), "%d")
all_trips$year <- format(as.Date(all_trips$date), "%Y")
all_trips$day_of_week <- format(as.Date(all_trips$date), "%A")
```

##### Added a "ride_length" calculation to all_trips (in seconds) and inspected the structure of my columns

```{r}
all_trips$ride_length <- difftime(all_trips$ended_at,all_trips$started_at)
```
```{r}
str(all_trips)
```

##### Converted ride_length from Factor to numeric so I can run calculations on the data

```{r}
is.factor(all_trips$ride_length)
all_trips$ride_length <- as.numeric(as.character(all_trips$ride_length))
is.numeric(all_trips$ride_length)


```

##### Removed "bad" data.The dataframe includes a few hundred entries when bikes were taken out of docks and checked for quality by Divvy or ride_length was negative will create a new version of the dataframe (v2) since data is being removed

```{r}
all_trips_v2 <- all_trips[!(all_trips$start_station_name == "HQ QR" | all_trips$ride_length<0),]
```

##### Conducted a descriptive analysis on my new clean Data Frame. I wanted to find out the average, median, max and min ride lengths

```{r}
mean(all_trips_v2$ride_length) #straight average (total ride length / rides)
median(all_trips_v2$ride_length) #midpoint number in the ascending array of ride lengths
max(all_trips_v2$ride_length) #longest ride
min(all_trips_v2$ride_length) #shortest ride

```
```{r}
summary(all_trips_v2$ride_length)
```

##### Compared members and casual users

```{r}
aggregate(all_trips_v2$ride_length ~ all_trips_v2$member_casual, FUN = mean)
aggregate(all_trips_v2$ride_length ~ all_trips_v2$member_casual, FUN = median)
aggregate(all_trips_v2$ride_length ~ all_trips_v2$member_casual, FUN = max)
aggregate(all_trips_v2$ride_length ~ all_trips_v2$member_casual, FUN = min)


```

##### average ride time by each day for members vs casual users. Fixed the days of the week so they are in order.

```{r}
aggregate(all_trips_v2$ride_length ~ all_trips_v2$member_casual + all_trips_v2$day_of_week, FUN = mean)
```

```{r}
all_trips_v2$day_of_week <- ordered(all_trips_v2$day_of_week, levels=c("Sunday", "Monday", "Tuesday", "Wednesday", "Thursday", "Friday", "Saturday"))
```

```{r}
aggregate(all_trips_v2$ride_length ~ all_trips_v2$member_casual + all_trips_v2$day_of_week, FUN = mean)
```

##### analyzed ridership data by type and weekday

```{r}
all_trips_v2 %>% mutate(weekdays = wday(started_at, label = TRUE)) %>% group_by(member_casual, weekdays) %>% summarise(number_of_rides = n(), average_duration = mean(ride_length)) %>% arrange(member_casual, weekdays)

```

### These visualization shows the number of rides by rider type and the average duration of each ride

```{r}
all_trips_v2 %>% 
  mutate(weekday = wday(started_at, label = TRUE)) %>% 
  group_by(member_casual, weekday) %>% 
  summarise(number_of_rides = n()
            ,average_duration = mean(ride_length)) %>% 
  arrange(member_casual, weekday)  %>% 
  ggplot(aes(x = weekday, y = number_of_rides, fill = member_casual)) +
  geom_col(position = "dodge")

```


```{r}
all_trips_v2 %>% 
  mutate(weekday = wday(started_at, label = TRUE)) %>% 
  group_by(member_casual, weekday) %>% 
  summarise(number_of_rides = n()
            ,average_duration = mean(ride_length)) %>% 
  arrange(member_casual, weekday)  %>% 
  ggplot(aes(x = weekday, y = average_duration, fill = member_casual)) +
  geom_col(position = "dodge")
```

##### A summary of my findings

```{r}
counts <- aggregate(all_trips_v2$ride_length ~ all_trips_v2$member_casual + all_trips_v2$day_of_week, FUN = mean)
write.csv(counts, file = "C:/Users/myola/Desktop/Divvy_Trips/avg_ride_length.csv")
```

## My findings

After going over the data about members and casual riders differ. There are a few observations that I have made. - On average a Casual rider will ride for 59min Vs a annual member average ride is for 14min - There seems to be a steady use from the annual members during the week and will decrease during the weekend. Where are casual riders will use the bike share more on the weekends. 
- On Average it seems like casual Riders spend more time when they do use the bike share service. And because of the pricing will end up paying more for one trip vs a annual plan.
- Cyclistic currently has a Annual Member plan that costs \$9/month for unlimited use for 45mins at a time. A Casual rider plan that cost $3.30 for each 30 min ride. Both charge $0.15/min after the allotted time.

## Recommendations

- Since we know that most people use the app to get a bike, we can have an alert that lets people know that their bike ride will almost be the same  price as an annual plan. Instead of paying just this ride they can pay it towards an annual membership.
- We can also send out an email to our exiting causl customers with an example of how much they've actually used the bike service and how much money they can save if they switch to an annual plan.
- There is a high chance to covert even just casual weekend rider because they will usually end up using the bike for much more than the 30min and will end up paying the same if not more.
